---
title: "Final Project: Citi Bikes Sample"
author: "Robert Berini"
format: html
editor: visual
execute: 
  warning: false
  message: false
---

## Introduction

The Citi Bike program is a bike-sharing system in New York City that allows users to rent bicycles for short trips around the city. The program has become increasingly popular since its launch in 2013, with millions of rides taken each year. In this project, we will analyze Citi Bike trip data for 2018 to gain insights into usage patterns, popular routes, and user profiles.

### Data Source

The data for this analysis is sourced from the Citi Bike system's public data repository, which provides monthly trip data in CSV format. The dataset includes information such as trip duration, start and end stations, user type (subscriber or customer), and timestamps for each trip.

### Data Dictionary

-   `tripduration`: Duration of the trip in seconds (numeric)
-   `tripduration_min`: Duration of the trip in minutes (numeric, calculated from tripduration)
-   `starttime`: Timestamp for when the trip started (datetime)
-   `stoptime`: Timestamp for when the trip ended (datetime)
-   `month`: Month the trip took place (ordered factor: Jan-Dec)
-   `day_of_week`: Day of the week the trip started (ordered factor: Mon-Sun)
-   `weekday_weekend`: Whether the trip occurred on a weekday or weekend (factor)
-   `hour_of_day`: Hour of day when the trip started, 0-23 (integer)
-   `time_of_day`: Time period classification (ordered factor: Morning, Late Morning, Afternoon, Evening, Night)
-   `start_station_id`: Unique identifier for the starting station (character)
-   `start_station_name`: Name of the starting station (character)
-   `start_station_latitude`: Latitude coordinate of the starting station (numeric)
-   `start_station_longitude`: Longitude coordinate of the starting station (numeric)
-   `end_station_id`: Unique identifier for the ending station (character)
-   `end_station_name`: Name of the ending station (character)
-   `end_station_latitude`: Latitude coordinate of the ending station (numeric)
-   `end_station_longitude`: Longitude coordinate of the ending station (numeric)
-   `round_trip`: Whether the trip started and ended at the same station (logical: TRUE/FALSE)
-   `bikeid`: Unique identifier for the bike used (character)
-   `usertype`: Type of user (factor: Subscriber or Customer)
-   `birth_year`: Self-reported birth year of the user (numeric)
-   `approx_age`: Approximate age of the user in 2018 (numeric, calculated as 2018 - birth_year)
-   `gender`: Self-reported gender of the user (factor: Male, Female, Unknown)

## Data Loading and Preparation

Load required packages

```{r}
library(conflicted)
library(tidyverse)
library(janitor)
library(urbnthemes)
library(broom)
library(skimr)
library(psych)
library(infer)
library(geomtextpath)
library(leaflet)

conflicts_prefer(
  dplyr::filter,
  urbnthemes::scale_fill_discrete,
  urbnthemes::scale_colour_discrete,
  urbnthemes::scale_fill_ordinal,
  janitor::chisq.test
)

options(
  scipen = 999,
  pillar.sigfig = 10
  )

set_urbn_defaults(style = "print")
```

Load the full transformed Citi Bike 2018 data (for reference)

```{r}
citibike_2018 <- readRDS("data/combined/citibike_2018_transformed.rds")
```

Load the Citi Bike 2018 data sample

```{r}
citibike_sample_2018 <-
  readRDS("data/sample/citibike_2018_sample_1000.rds")

glimpse(citibike_sample_2018)
```

## Exploratory data analysis

### Summary statistics

```{r}
citibike_sample_2018 |>
  summary()
```

```{r}
citibike_sample_2018 |>
  skim()
```

### Trip duration analysis

```{r}
citibike_sample_2018 |>
  select(tripduration_min) |>
  describe()
```

```{r}
citibike_sample_2018 |>
  ggplot(aes(x = tripduration_min)) +
  geom_density() +
  labs(
    title = "Distribution of Trip Duration",
    x = "Trip Duration (minutes)",
    y = "Density"
  ) +
  remove_axis(axis = "y") +
  remove_ticks()
```

```{r}
citibike_sample_2018 |>
  ggplot(aes(x = tripduration_min)) +
  geom_density() +
  labs(
    title = "Distribution of Trip Duration",
    x = "Trip Duration (minutes)",
    y = "Density"
  ) +
  xlim(0, 60) +
  remove_axis(axis = "y") +
  remove_ticks()
```

### Bootstrap confidence interval for mean trip duration

```{r}
set.seed(557)

citibike_sample_2018 |>
  slice_sample(n = nrow(citibike_sample_2018), replace = TRUE) |>
  pull(tripduration_min) |>
  mean(na.rm = TRUE)
```

```{r}
set.seed(557)

boot_dist_1 <- citibike_sample_2018 |>
  specify(response = tripduration_min) |>
  generate(reps = 1, type = "bootstrap") |>
  calculate(stat = "mean")

print(boot_dist_1)

boot_dist_1 |>
  ggplot(aes(x = stat)) +
  geom_dotplot(binwidth = 0.1, fill = "#1696d2", stroke = 0) +
  xlim(10, 16) +
  labs(x = "Bootstrap Mean Trip Duration", y = "") +
  theme(axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.minor = element_blank()
      )
```

```{r}
set.seed(557)

boot_dist_10 <- citibike_sample_2018 |>
  specify(response = tripduration_min) |>
  generate(reps = 10, type = "bootstrap") |>
  calculate(stat = "mean")

print(boot_dist_10)

boot_dist_10 |>
  ggplot(aes(x = stat)) +
  geom_dotplot(binwidth = 0.1, fill = "#1696d2", stroke = 0) +
  xlim(10, 16) +
  labs(x = "Bootstrap Mean Trip Duration", y = "") +
  theme(axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.minor = element_blank()
      )
```

```{r}
set.seed(557)

boot_dist_1000 <- citibike_sample_2018 |>
  specify(response = tripduration_min) |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "mean")

boot_dist_1000 |>
  ggplot(aes(x = stat)) +
  geom_dotplot(binwidth = 0.1, dotsize = .45, fill = "#1696d2", stroke = 0) +
  xlim(10, 16) +
  labs(x = "Bootstrap Mean Trip Duration", y = "") +
  theme(axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.minor = element_blank()
      )
```

```{r}
boot_ci <- boot_dist_1000 |>
  get_confidence_interval(level = 0.95, type = "percentile")

print(boot_ci)

boot_dist_1000 |>
  visualize(bins = 100, fill = "#1696d2", stroke = 0) +
  shade_confidence_interval(endpoints = boot_ci, fill = "#fdbf11", color = "#fdbf11") +
  xlim(10, 16) +
  labs(x = "Bootstrap Mean Trip Duration", y = "") +
  theme(axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.minor = element_blank()
      )
```

### Mathematical confidence interval for mean trip duration

```{r}
citibike_sample_2018 |>
  select(tripduration_min) |>
  t.test(conf.level = 0.95) |>
  tidy()
```

### Comparison of bootstrap and mathematical confidence intervals

```{r}
pop_mean <- mean(citibike_2018$tripduration_min, na.rm = TRUE)

boot_mean <- mean(boot_dist_1000$stat)

t_result <- citibike_sample_2018 |>
  select(tripduration_min) |>
  t.test(conf.level = 0.95) |>
  tidy()

tibble(
  Method = c("Population (True)", "Bootstrap", "T-test"),
  Mean = c(pop_mean, boot_mean, t_result$estimate),
  Lower_CI = c(NA, boot_ci$lower_ci, t_result$conf.low),
  Upper_CI = c(NA, boot_ci$upper_ci, t_result$conf.high)
)
```

### Bivariate analysis: Trip duration by weekday vs. weekend

```{r}
citibike_sample_2018 |>
  ggplot(aes(x = tripduration_min, fill = weekday_weekend)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Distribution of Trip Durations by Week Part",
    x = "Trip Duration (minutes)",
    y = "Density",
    fill = "Week Part"
  ) +
  xlim(0, 60) +
  remove_axis(axis = "y") +
  remove_ticks()
```

```{r}
citibike_sample_2018 |>
  ggplot(aes(x = weekday_weekend, y = tripduration_min, fill = weekday_weekend)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
  stat_summary(
    fun = mean, 
    geom = "point", 
    shape = 23,
    size = 4, 
    fill = "#db2b27",
    color = "#000000",
    show.legend = FALSE
    ) +
  labs(
    title = "Trip Duration Distribution by Week Part",
    x = "",
    y = "Trip Duration (minutes)"
  ) +
  coord_flip(ylim = c(0, 30))
```

### Randomization test: Mean trip duration on weekdays vs. weekends

Calculate observed difference in means

```{r}
obs_diff <- citibike_sample_2018 |>
  specify(tripduration_min ~ weekday_weekend) |>
  calculate(stat = "diff in means", order = c("Weekend", "Weekday"))

obs_diff
```

Generate null distribution through permutation

```{r}
set.seed(557)

null_distribution <- citibike_sample_2018 |>
  specify(tripduration_min ~ weekday_weekend) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in means", order = c("Weekend", "Weekday"))
```

Visualize null distribution with observed statistic

```{r}
null_distribution |>
  visualize() +
  shade_p_value(obs_stat = obs_diff$stat, direction = "two_sided") +
  labs(
    title = "Null Distribution of Mean Trip Duration Difference",
    x = "Difference in Mean Trip Duration (Weekend - Weekday)",
    y = "Frequency"
  ) +
  remove_axis(axis = "y") +
  remove_ticks()
```

```{r}
null_distribution |>
  get_p_value(obs_stat = obs_diff$stat, direction = "two-sided")
```

### Mathematical two-sample t-test: Mean trip duration on weekdays vs. weekends

```{r}
citibike_sample_2018 |>
  mutate(weekday_weekend = fct_relevel(weekday_weekend, "Weekend", "Weekday")) |>
  (\(x) t.test(tripduration_min ~ weekday_weekend, data = x, var.equal = FALSE))() |>
  tidy()
```

### Comparison of randomization test and t-test results

```{r}
true_diff <- citibike_2018 |>
  group_by(weekday_weekend) |>
  summarize(mean_duration = mean(tripduration_min, na.rm = TRUE)) |>
  summarize(diff = diff(mean_duration)) |>
  pull(diff)

randomization_result <-
  null_distribution |>
  get_p_value(obs_stat = obs_diff$stat, direction = "two-sided") |>
  pull(p_value)

t_test_result <- citibike_sample_2018 |>
  mutate(weekday_weekend = fct_relevel(weekday_weekend, "Weekend", "Weekday")) |>
  (\(x) t.test(tripduration_min ~ weekday_weekend, data = x, var.equal = FALSE))() |>
  tidy()

tibble(
  Method = c("Population (True)", "Randomization Test", "T-test"),
  Difference = c(
    true_diff,
    obs_diff$stat,
    t_test_result$estimate
  ),
  P_Value = c(
    NA,
    randomization_result,
    t_test_result$p.value
  )
)
```

### Bivariate analysis: Trip duration by user type

```{r}
citibike_sample_2018 |>
  ggplot(aes(x = tripduration_min, fill = usertype)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Distribution of Trip Durations by User Type",
    x = "Trip Duration (minutes)",
    y = "Density",
    fill = "User Type"
  ) +
  xlim(0, 60) +
  remove_axis(axis = "y") +
  remove_ticks()
```

```{r}
citibike_sample_2018 |>
  ggplot(aes(x = usertype, y = tripduration_min, fill = usertype)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
  stat_summary(
    fun = mean, 
    geom = "point", 
    shape = 23,
    size = 4, 
    fill = "#db2b27",
    color = "#000000",
    show.legend = FALSE
    ) +
  labs(
    title = "Trip Duration Distribution by Usertype",
    x = "",
    y = "Trip Duration (minutes)"
  ) +
  coord_flip(ylim = c(0, 60))
```

```{r}
citibike_sample_2018 |>
  group_by(usertype) |>
  summarize(median_duration = median(tripduration_min, na.rm = TRUE)) |>
  ggplot(aes(x = usertype, y = median_duration, fill = usertype)) +
  urbnthemes::geom_col(show.legend = FALSE) +
  labs(
    title = "Median Trip Duration by User Type",
    x = "",
    y = "Median Trip Duration (minutes)"
  ) +
  coord_flip()
```

```{r}
obs_diff_median <- citibike_sample_2018 |>
  specify(tripduration_min ~ usertype) |>
  calculate(stat = "diff in medians", order = c("Customer", "Subscriber"))

obs_diff_median
```

```{r}
set.seed(557)

null_distribution_median <- citibike_sample_2018 |>
  specify(tripduration_min ~ usertype) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in medians", order = c("Customer", "Subscriber"))
```

```{r}
null_distribution_median |>
  visualize() +
  shade_p_value(obs_stat = obs_diff_median, direction = "two-sided") +
  labs(
    title = "Null Distribution of Difference in Median Trip Duration",
    x = "Difference in Median Duration (minutes)",
    y = "Count"
  )
```

```{r}
null_distribution_median |>
  get_p_value(obs_stat = obs_diff_median, direction = "two-sided")
```

```{r}
wilcox.test(tripduration_min ~ usertype,
            data = citibike_sample_2018,
            exact = FALSE) |>
  tidy()
```

```{r}
true_diff_median <- citibike_2018 |>
  group_by(usertype) |>
  summarize(median_duration = median(tripduration_min, na.rm = TRUE)) |>
  summarize(diff = diff(median_duration)) |>
  pull(diff)

randomization_result_median <-
  null_distribution_median |>
  get_p_value(obs_stat = obs_diff_median, direction = "two-sided") |>
  pull(p_value)

wilcox_result <- wilcox.test(tripduration_min ~ usertype,
                              data = citibike_sample_2018,
                              exact = FALSE) |>
  tidy()

tibble(
  Method = c("Population (True)", "Randomization Test", "Wilcoxon Test"),
  Difference = c(
    true_diff_median,
    obs_diff_median$stat,
    obs_diff_median$stat  # Same observed difference
  ),
  P_Value = c(
    NA,
    randomization_result_median,
    wilcox_result$p.value
  )
)
```

```{r}
citibike_sample_2018 |>
  ggplot(aes(x = time_of_day, y = tripduration_min, fill = time_of_day)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
  stat_summary(
    fun = mean, 
    geom = "point", 
    shape = 23,
    size = 3, 
    fill = "#db2b27",
    color = "#000000",
    show.legend = FALSE
    ) +
  labs(
    title = "Trip Duration Distribution by Time of Day",
    x = "",
    y = "Trip Duration (minutes)"
  ) +
  coord_flip(ylim = c(0, 30))
```

```{r}
aov(tripduration_min ~ time_of_day, data = citibike_sample_2018) |>
  tidy()
```

```{r}
citibike_sample_2018 |>
  count(hour_of_day, usertype) |>
  ggplot(aes(x = hour_of_day, y = n, color = usertype)) +
  geom_line(linewidth = 1) +
  geom_label(
    data = ~group_by(.x, usertype) |> filter(hour_of_day == max(hour_of_day)),
    aes(label = usertype),
    hjust = -0.1,
    size = 3,
    fontface = "bold",
    fill = "white",
    label.size = 0,
    label.padding = unit(0.15, "lines")
  ) +
  scale_x_continuous(
    breaks = seq(0, 24, by = 6),
    expand = expansion(mult = c(0.02, 0.15))
  ) +
  labs(title = "Trip Volume by Hour and User Type",
       x = "Hour of Day", y = "Number of Trips") +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off")
```

```{r}
citibike_sample_2018 |>
  count(month, usertype) |>
  ggplot(aes(x = month, y = n, color = usertype, group = usertype)) +
  geom_line(linewidth = 1) +
  geom_label(
    data = ~group_by(.x, usertype) |> filter(month == max(month)),
    aes(label = usertype),
    hjust = -0.1,
    size = 3,
    fontface = "bold",
    fill = "white",
    label.size = 0,
    label.padding = unit(0.15, "lines")
  ) +
  scale_x_discrete(
    expand = expansion(mult = c(0.02, 0.15))
  ) +
  labs(title = "Trip Volume by Month and User Type",
       x = "Month", y = "Number of Trips") +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off")
```

```{r}
citibike_sample_2018 |>
  ggplot(aes(x = approx_age, y = tripduration_min)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Trip Duration by Age",
    x = "Age (years)",
    y = "Trip Duration (minutes)"
  ) +
  scatter_grid()
```

```{r}
citibike_sample_2018 |>
  filter(tripduration_min <= 60) |>
  ggplot(aes(x = approx_age, y = tripduration_min)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Trip Duration by Age (trips ≤ 60 minutes)",
    x = "Age (years)",
    y = "Trip Duration (minutes)"
  ) +
  scatter_grid()
```

```{r}
cor.test(citibike_sample_2018$approx_age, 
         citibike_sample_2018$tripduration_min) |>
  tidy()
```

```{r}
citibike_sample_2018 |>
  group_by(usertype, gender) |>
  summarize(
    n = n(),
    mean_age = mean(approx_age, na.rm = TRUE),
    median_age = median(approx_age, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
chisq.test(citibike_sample_2018$usertype, citibike_sample_2018$gender) |>
  tidy()
```

### Popular start stations

```{r}
citibike_sample_2018 |>
  count(start_station_name, sort = TRUE) |>
  slice_head(n = 10) |>
  ggplot(aes(x = reorder(start_station_name, n), y = n)) +
  urbnthemes::geom_col() +
  coord_flip() +
  labs(title = "Top 10 Most Popular Start Stations",
       x = "", y = "Number of Trips")
```

### Map of start stations by trip volume

```{r}
station_counts <- citibike_sample_2018 |>
  count(start_station_name, start_station_latitude, start_station_longitude) |>
  mutate(
    top_station = if_else(rank(desc(n)) <= 10, "Top 10", "Other"),
    color = if_else(top_station == "Top 10", "#0a4c6a", "#ca5800")
  ) |>
  arrange(top_station)

leaflet(station_counts) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~start_station_longitude,
    lat = ~start_station_latitude,
    radius = ~sqrt(n) * 2,
    popup = ~paste0(start_station_name, "<br>Trips: ", n),
    fillOpacity = 0.7,
    color = ~color,
    fillColor = ~color,
    stroke = TRUE,
    weight = 1
  ) |>
  addLegend(
    position = "bottomright",
    colors = c("#0a4c6a", "#ca5800"),
    labels = c("Top 10", "Other"),
    title = "Start Station Volume"
  )
```

### Popular end stations

```{r}
citibike_sample_2018 |>
  count(end_station_name, sort = TRUE) |>
  slice_head(n = 10) |>
  ggplot(aes(x = reorder(end_station_name, n), y = n)) +
  urbnthemes::geom_col() +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, 10, by = 2.5)) +
  labs(title = "Top 10 Most Popular Destination Stations",
       x = "", y = "Number of Trips")
```

### Map of popular start and end stations

```{r}
start_counts <- citibike_sample_2018 |>
  count(start_station_name, start_station_latitude, start_station_longitude) |>
  slice_max(n, n = 10) |>
  mutate(type = "Top Start")

end_counts <- citibike_sample_2018 |>
  count(end_station_name, end_station_latitude, end_station_longitude) |>
  rename(
    start_station_name = end_station_name,
    start_station_latitude = end_station_latitude,
    start_station_longitude = end_station_longitude
  ) |>
  slice_max(n, n = 10) |>
  mutate(type = "Top End")

station_counts <- bind_rows(end_counts, start_counts) |>
  mutate(
    color = if_else(type == "Top Start", "#0a4c6a", "#ca5800")
  )

leaflet(station_counts) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~start_station_longitude,
    lat = ~start_station_latitude,
    radius = ~sqrt(n) * 2.5,
    popup = ~paste0(start_station_name, "<br>", type, "<br>Trips: ", n),
    fillOpacity = 0.7,
    color = ~color,
    fillColor = ~color,
    stroke = TRUE,
    weight = 1
  ) |>
  addLegend(
    position = "bottomright",
    colors = c("#0a4c6a", "#ca5800"),
    labels = c("Top 10 Start Stations", "Top 10 End Stations"),
    title = "Station Type"
  )
```
